\propertitle

### 2019-04-20 maxNt_v5.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-04-19T19:24:02+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal population\\ from subpopulation constraints%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample constraints}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

%\DeclarePairedDelimiterX{\condt}[2]{[}{]}{#1\nonscript\,\delimsize\vert\nonscript\;\mathopen{}#2}

@@

\section{Maximum-entropy and recordings of neuronal activity}
\label{sec:intro}

A necessary step of every plausible inference is the assignment of some
probabilistic premisses -- as many as necessary to arrive at our desired
probabilistic conclusions given the evidence. The necessity of this step
was tersely stated by W.~E.~Johnson %(J.~M.~Keynes's supervisor)
a hundred years ago:\vspace{\topsep}
\begin{quote}\footnotesize
  Now the axioms of probability enable us to infer any
  probability-conclusion \emph{only} from probability-premisses. In other
  words, the calculus of probability does not enable us to infer any
  probability-value unless we have some probabilities or probability
  relations \emph{given}. Such data cannot be supplied by the
  mathematician. E.g. the rules of arithmetic and the axioms of the
  probability-calculus are utterly impotent to determine, on the supposed
  knowledge that the throw of a coin must yield either head or tail and
  cannot yield both, the probability that it will yield head or that it
  will yield tail. We must assume that the two co-exclusive and
  co-exhaustive possibilities are \emph{equally probable}, before we can
  estimate the probability of either as being a half of certitude.
  \sourceatright{\citep[Appendix on eduction, \sect~5,
    p.~182]{johnson1924}}
\end{quote}

@@

We can for example assign probabilistic premisses at a very deep
level, as schematically illustrated on the left. They determine the
probabilities of statements at all higher levels, including the statements
we're ultimately interested in.

\begin{wrapfigure}{r}{0.33\linewidth}% with wrapfigure
\centering\includegraphics[width=\linewidth]{statements_scheme2b.pdf}%\\
%\caption{caption}\label{fig:comparison_a5}
\end{wrapfigure}% statements_scheme.svg statements_scheme2.svg
Or we can assign probabilistic premisses at a level of intermediate depth,
as schematically illustrated on the right. In this case they determine the
probabilities of statements at all higher levels, and partly constrain
the probabilities of statements at deeper levels.

The choice of the logical level at which to assign the premisses depends on
many factors: on the difficulty of assessing our state of knowledge at
particular levels, on the cost of the ensuing calculations, and especially
on the general purpose of our inference. But two assignments of premisses at
different levels can still turn out to be completely equivalent for the
probabilities of the statements of interest.

Here's an example. Suppose that $n$ neurons have been recorded in a
sequence of time bins. We're given the time averages of the numbers of
active neurons and neuron pairs in the sequence. We don't know the exact
sequence of activities or the number $T$ of time bins, although we know
that the latter is large. We want to forecast how many neurons will be
active in a \emph{new} time bin.

In a deep approach to this inference problem we assign the probability for
the possible value of $T$, and the probabilities for several hypotheses
about the sequence of activities in $T+1$ or more time bins. One hypothesis
could be, for example, that the $n$ neurons are a sample of a larger
population of $N$ neurons and for this population there are some sufficient
statistics, out of biological reasons. We could of course deepen this
approach further, with no end, down to assigning the probabilities for the
constitution and motion of the molecules of the $N$ neurons and their
environment, the probabilities for the position and response of the
recording electrodes, and so on. From all these probabilities and the data
we could finally calculate the probability for the activity in the new time
bin.

At the other end, the shallowest approach to this inference problem is to
directly assign the probabilities for the number of active neurons in the
new time bin, somehow using the data we were given. (It must be remarked
that such assignment partly constrains, in a backward fashion, all assignments we
would do at a deeper level.)

The maximum-entropy method was proposed by Jaynes
\parentext{\cite*{jaynes1957,jaynes1963,jaynes1994_r2003};
  \cite{sivia1996_r2006,hobsonetal1973}} as a way of assigning probability
premisses at a more direct level, as in the approach just described. This
point is made especially clear in \textcite{jaynes1994_r2003},
\sect~11.1:\vspace{0.5\topsep}
\begin{quote}\footnotesize
  If we knew the numbers $N$, $L$, $W$, then this could be solved by direct
  application of Bayes' theorem; without that information, we could still
  introduce the unknowns $N$, $L$, $W$ as nuisance parameters and use
  Bayes' theorem, eliminating them at the end. \textelp{} However, the
  Bayesian solution would not really address our problem; it only transfers
  it to the problem of assigning priors to $N$, $L$, $W$, leaving us back
  in essentially the same situation; how do we assign informative
  probabilities?
\end{quote}
The idea behind this method is to translate \emph{all} the data we're
given into constraints -- such as expectations -- for our probability
assignment, resolving the leftover freedom by choosing the constrained
probability distribution having highest Shannon entropy, as a measure of
its uncertainty or broadness. It can be shown that the use of this
method is approximately equivalent, under specific conditions, to a
probability assignment at a deeper level based on an assumption of
exchangeability together with a multinomial probability distribution for
long-run frequencies, or with an assumption about sufficient statistics
\citep{jaynes1986d_r1996,rodriguez1989,rodriguez2002,skilling1989b,skilling1990}[\sect~4.5]{bernardoetal1994_r2000}{diaconisetal1981,portamana2017,portamana2009}.

@@

In the present work we show how to use the maximum-entropy method to
assign a probability distribution for the activity of a population of
neurons at a new time bin, given specific statistics about the activity of
a \emph{subpopulation}, that is, a sample, of the full population. This
kind of probability assignment is extremely important in neuroscience: we
try to understand the workings of brain areas that may include tens of
thousands of neurons, but we can record only few tens or hundreds
of them.

Our approach also corrects a slight misuse of the maximum-entropy method as
it has appeared in the literature. When this method is applied to a set of
neurons, it implicitly assumes that the set is completely isolated from
other neurons. We'll reveal this implicit assumption in several ways. The
probability distribution thus constructed is therefore affected by this
hidden assumption. We'll study in which circumstances the application of
the method at the sample level gives results similar to the application at
the full-population level.

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-04-20T15:37:10+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal population\\ from subpopulation constraints%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample constraints}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

\DeclarePairedDelimiterX{\condt}[2]{[}{]}{#1\nonscript\,\delimsize\vert\nonscript\;\mathopen{}#2}

@@

\section{Why polls?}
\label{sec:polls}

When political elections are approaching, mass media are constantly
bubbling with results and discussions of pre-election polls. Why do we
conduct pre-election polls and other kinds of survey sampling? The
simplest, intuitive answer is \enquote{because they give us an idea of what
  the final election outcomes will be}. Can this answer be made more
quantitative?

Yes. In the simplest cases, the probability calculus give us simple and
precise quantitative relations between our expectations from a population
-- of voters, for instance -- and from a sample thereof. Suppose there are
$N$ voters, $S$ of which prefer one political candidate and $N-S$ of which
prefer another. We don't know $S$ or $N-S$. Now take a sample of $n$
voters, assuming that every set of $n$ voters is equally likely to be
sampled. In this sample, $s$ voters prefer the first candidate, and $n-s$
the second. The probability calculus tells us that our expectations of
$s/n$ and $S/N$ are equal:
\begin{equation}
  \label{eq:equal_expectations_poll}
  \E(s/n \| \yH) = \E(S/N \| \yH),
\end{equation}
where $\yH$ represents our state of knowledge in this example.

A more precise answer to our original question is then: \enquote{because we
  \emph{expect} to see the same proportion of political preferences in a
  sample as in the full population}.

The equality above extends to other quantities beyond the means $S/N$ and
$s/n$: it holds for all normalized \emph{factorial} moments \citep{potts1953,broca2005}:
\begin{multline}
  \label{eq:equal_factorialmoments_polls}
  \E\condt[\Big]{\tfrac{s\;(s-1)}{n\;(n-1)}}{\yH} = 
  \E\condt[\Big]{\tfrac{S\;(S-1)}{N\;(N-1)}}{\yH},\\
  \E\condt[\Big]{\tfrac{s\;(s-1)\;(s-2)}{n\;(n-1)\;(n-2)}}{\yH} = 
  \E\condt[\Big]{\tfrac{S\;(S-1)\;(S-2)}{N\;(N-1)\;(N-2)}}{\yH},\quad\dotsc
  \\
  \E\condt[\Big]{\tbinom{s}{r}\;\tbinom{n}{r}^{-1}}{\yH} =
  \E\condt[\Big]{\tbinom{S}{r}\;\tbinom{N}{r}^{-1}}{\yH},
\end{multline}
with $r<N$.

\section{Maximum-entropy method}
\label{sec:maxent_remarks}

A necessary step of every plausible inference is the assignment of some
probabilistic premisses; as many as necessary to arrive at our desired
probabilistic conclusions given the evidence. The necessity of this step
was tersely stated by W.~E.~Johnson %(J.~M.~Keynes's supervisor)
a century ago:\vspace{\topsep}
\begin{quote}\footnotesize
  Now the axioms of probability enable us to infer any
  probability-conclusion \emph{only} from probability-premisses. In other
  words, the calculus of probability does not enable us to infer any
  probability-value unless we have some probabilities or probability
  relations \emph{given}. Such data cannot be supplied by the
  mathematician. E.g. the rules of arithmetic and the axioms of the
  probability-calculus are utterly impotent to determine, on the supposed
  knowledge that the throw of a coin must yield either head or tail and
  cannot yield both, the probability that it will yield head or that it
  will yield tail. We must assume that the two co-exclusive and
  co-exhaustive possibilities are \emph{equally probable}, before we can
  estimate the probability of either as being a half of certitude.
  \sourceatright{\citep[Appendix on eduction, \sect~5,
    p.~182]{johnson1924}}
\end{quote}

@@

We can for example assign probabilistic premisses at a very deep analytical
level, as schematically illustrated on the left. They determine the
probabilities of statements at more superficial levels, including the
statements which we're ultimately interested in or which we can directly
observe.

\begin{wrapfigure}{r}{0.33\linewidth}% with wrapfigure
\centering\includegraphics[width=\linewidth]{statements_scheme2b.pdf}%\\
%\caption{caption}\label{fig:comparison_a5}
\end{wrapfigure}% statements_scheme.svg statements_scheme2.svg
Or we can assign probabilistic premisses at a level of superficial or
intermediate depth, as schematically illustrated on the right. In this case
they determine the probabilities of statements at more superficial levels, and
partly constrain the probabilities of statements at deeper levels.

The choice of the analytical depth at which to assign the premisses depends
on many factors: on the difficulty of assessing our state of uncertainty at
that depth, on the cost of the ensuing calculations, and especially on the
general purpose of our inference. Assignments of probabilistic premisses at
different levels can still turn out to be completely equivalent for the
probabilities of the statements of interest.

Here's an example. Suppose that $n$ neurons have been recorded in a
sequence of time bins. We're given the time averages of the numbers of
active neurons and of neuron pairs in the sequence. We don't know the exact
sequence of activities or the number $T$ of time bins, although we know
that the latter is large. We want to forecast how many of the $n$ neurons
will be active in a \emph{new} time bin.

In a deep approach to this inference problem we assign the probability for
the possible value of $T$, and the probabilities for several hypotheses
about the sequence of activities in $T+1$ or more time bins. One hypothesis
could be, for instance, that the $n$ neurons are a sample of a larger
population of $N$ neurons, and a second hypothesis could be that for this
population there are some sufficient statistics, out of biological reasons.
We could of course deepen this approach further, with no end, down to
assigning the probabilities for the constitution and motion of the
molecules of the neurons and their environment, the probabilities for the
position and response of the recording electrodes, and so on. From all
these probabilities and the data we could finally calculate the probability
for the activity in the new time bin.

At the other end, the shallowest approach to this inference problem is to
directly assign the probabilities for the number of active neurons in the
new time bin, somehow using the data we were given. (It must be remarked
that such assignment partly constrains, in a backward fashion, all
assignments we could do at a deeper level.)

The maximum-entropy method was proposed by Jaynes
\parentext{\cite*{jaynes1957,jaynes1963,jaynes1994_r2003};
  \cite{sivia1996_r2006,hobsonetal1973}} as a way of assigning probability
premisses at a more direct level, as in the approach just described. This
point is made especially clear in \textcite{jaynes1994_r2003},
\sect~11.1:\vspace{0.5\topsep}
\begin{quote}\footnotesize
  If we knew the numbers $N$, $L$, $W$, \textins{some quantities relevant
    to our inference} then this could be solved by direct application of
  Bayes' theorem; without that information, we could still introduce the
  unknowns $N$, $L$, $W$ as nuisance parameters and use Bayes' theorem,
  eliminating them at the end. \textelp{} However, the Bayesian solution
  would not really address our problem; it only transfers it to the problem
  of assigning priors to $N$, $L$, $W$, leaving us back in essentially the
  same situation; how do we assign informative probabilities?
\end{quote}
The idea behind this method is to translate \emph{all} the data we're
given into constraints -- such as expectations -- for our probability
assignment, resolving the leftover freedom by choosing the constrained
probability distribution having highest Shannon entropy, as a measure of
its uncertainty or broadness. It can be shown that the use of this
method is approximately equivalent, under specific conditions, to a
probability assignment at a deeper level based on an assumption of
exchangeability together with a multinomial probability distribution for
long-run frequencies, or with an assumption about sufficient statistics
\citep{jaynes1986d_r1996,rodriguez1989,rodriguez2002,skilling1989b,skilling1990}[\sect~4.5]{bernardoetal1994_r2000}{diaconisetal1981,portamana2017,portamana2009}.

@@

In the present work we show how to use the maximum-entropy method to assign
a probability distribution for the total activity at a new time bin of a
population of neurons, \emph{most of which can be unrecorded}, given
specific statistics about the activity of a recorded \emph{sample} of the
full population. This kind of probability assignment is extremely important
in neuroscience: we try to understand the workings of brain areas that may
include tens of thousands of neurons, but we can record only few tens or
hundreds of them.

Our approach also corrects a slight misuse of the maximum-entropy method
as sometimes appears in the literature. When this method is applied to a
set of neurons, it implicitly assumes that the set is completely isolated
from other neurons. We'll reveal this implicit assumption in several ways.
The probability distribution thus constructed is therefore affected by this
hidden assumption. We'll study in which circumstances the application of
the method at the sample level and at the full-population level give
similar or different probabilities.

The maximum-entropy method can be used this way thanks to a very simple ***
